{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Predictions for BHBH, BHNS and NSNS binaries in LISA

### 2021-11-02 paper/tex/sections/variations.tex
proof read variations

The efficiency of mass transfer, i.e.\ the fraction of mass lost by the donor through Roche lobe overflow that is accreted by the companion, is poorly constrained and is considered as one of the main uncertainties in binary evolution \citep[e.g.\ ][]{deMink+2007}. In our fiducial model A, we use a prescription where the 
accretion rate onto stellar companions is regulated by their thermal timescale, i.e. the timescale on which a star can react to changes and restore thermal equilibrium \citep[see e.g.\ ][]{Schneider+2015}.

In models \modBetaLow{}-\modBetaHigh{}, we adopt a fixed value for the mass transfer efficiency, $\beta$, from $\beta=0.25$ up to 0.75, in cases of stable mass transfer onto a stellar companion. For accretion onto NS and BH we still assume that their accretion is limited to the Eddington rate.

@@

For the most massive progenitors, increasing $\beta$ leads to secondary stars that are so massive and luminous that they experience strong wind mass loss. This leads to further widening of the orbit and it limits the amount by which the secondaries expand. Both of these effects can prevent secondaries from filling their Roche lobe. This means that they cannot initiate the reverse interaction needed to shrink the binary system and eventually produce a detectable compact binary system. We indeed see that increasing the mass transfer efficiency leads to 
a decrease of the expected number of detections for BHBHs and BHNSs, which originate from the most massive progenitors.

For lower mass progenitors, which primarily produce NSNS systems, we find the opposite. Increasing $\beta$ leads to an increase of detectable systems. This is in part because the increased mass of the secondary stars allows stars that would have other wise ended their life as a WD, can now become massive enough to form a NS. The same effect also allows some NS progenitors to become massive enough to BH progenitors, but this effect is less strong. Furthermore, the changes in secondary mass and orbital widening, affects the number of systems for which the reverse interaction successfully ejects the envelope and shrinks the orbit. Indeed, the net effect is an increase of NSNS systems with increasing $\beta$.

@@

The common-envelope phase constitutes the most challenging phase in the evolution of interacting binary systems \citep{Ivanova+2013}. The debate concerns the conditions required for the onset of a common envelope phase and, if a common envelope phase occurs, what the outcome is. Rapid populations synthesis simulations such as ours approximate both questions in a crude way. We therefore consider several model variations.

\paragraph{The efficiency parameter $\alpha_{\rm CE}$} To estimate the outcome of a CE phase, we use a simple consideration of the binding energy and orbital energy  \citep{Webbink+1984}. Our fiducial model assumes a common-envelope efficiency parameter $\alpha_{\rm CE} = 1$ which can be interpreted as the case where all the energy liberated by shrinking the orbit it used in an optimal way to unbind the envelope \citep{Webbink+1984}. There have been many attempts to constrain this parameter using observations and more recently also using 3D simulations, but no consistent picture has emerged. Therefore we consider variations in this parameter.

@@

Increasing $\alpha_{\rm CE}$ makes common-envelope ejection more efficient or, in other words, less orbital shrinkage is needed to successfully eject the envelope. This has two consequences. (1) Firstly, a larger fraction of systems avoids merging during the CE phase. This increases the overall number of double compact object binaries. (2) Secondly, the systems that survive the CE phase are wider, possibly too wide to become detectable as gravitational wave sources. So, while the first consequence favours the formation of double compact object systems, the second consequence disfavours the subset of those close enough to be detected.

These two opposing effect result in fine tuning situation. Only a very small subset of progenitor systems that have the right orbital parameters prior to the CE phase to successfully produce detectable systems. Changing  $\alpha_{\rm CE}$ moves and changes this window in the parameter space that successfully leads to the formation of detectable systems. How the number of the detectable systems changes depends on whether the relevant parameter space grows or shrinks and on how well the relevant part of the parameters space is populated. Fully unravelling these effects and how they interplay with the assumed star formation history is beyond our scope (and possibly not even of large relevance given the severe simplifications). We will limit the further discussions to simply stating the trends we observe.

@@

\paragraph{The ``optimistic'' CE treatment} We further explore a model variation introduced by \citet{Belczynski+2007} often referred to as the ``optimistic'' CE scenario. This model also considers donor stars that are on the Hertzsprung as the progenitors for CE events.

@@

\paragraph{Case BB mass transfer} For completeness and consistency with \citet{Broekgaarden+2021}, we have also included model variations that consider uncertainties in case BB mass transfer. his is a phase of mass transfer where the donor star has already lost its hydrogen envelope in a prior interaction, but fills its Roche lobe again as expands during helium shell burning phase \citep[e.g.][]{Dewi+2003, Vigna-Gomez+2018}. This is of particular interest for the formation of NS stars, as their lower mass progenitors are swelling the most during this phase \citet[e.g.][and references therein]{Laplace+2020}. Nearly all NSNS systems from through case BB mass transfer

In model \modCaseBB{} we effectively exclude any systems that evolve through case BB mass transfer. As expected, this kills nearly all NSNS systems. We also find a heavy reduction of the BHNS systems y a factor 6. The BHBH systems are not affected, as expected, since case BB mass transfer does not play a role for high mass progenitors.

Finally, in model {[\bf TOM.. please add label]} [=F] we consider a situation where case BB mass transfer always leads to unstable mass transfer and initiates a CE phase combined with the optimistic treatment for CE. This model leads to the highest predictions for the detection's among all variations that we have considered.

@@

The formation of compact remnants and and their associated natal kicks also constitute important uncertainties.

In model \modRapid{} we consider the so-called ``rapid'' remnant mass function by \citet{Fryer+2015} as an alternative to the ``delayed'' description used in our detailed simulations. This affects the mass distribution (see Section~\ref{XX}), but the effects on the number of detectable systems is modest for BHBH and BHNS, and negligible for NSNS.

The same is true for changing the impact of changing the assumed maximum neutron star mass, $m_{\rm ns}$ (\modNSLow{} and \modNSHigh{}). Lowering $m_{\rm ns}$  increases the number of detections involving BHs and vice versa, but has no significant effect on the number of NSNS detections.

In Model \modNoPISN{}, we ignore pair instability supernovae, which has no significant effect on the predictions for the number of detections with LISA. This is not a surprise as the typical metallicity of the progenitors in our Milky Way is rather high.  The detectable systems are dominated by systems with final core masses that are well below the mass regime for pair instability supernovae.

Decreasing the natal kicks for all core-collapse supernovae (models \modSigLow{}-\modSigLower{}) increases the detection rates for each DCO type since lower kicks result in fewer disrupted binaries and hence a more numerous detectable population. The NSNS systems are most strongly affected.   The impact on BHNS systems is mild and even insignificant for BHBH systems. This is because, in our models, the natal kicks for BHs are scaled down with the amount of mass that falls back, which is large in the case of BHs.

In model \modNoBH{} we assume BHs form without any kick, while using our fiducial assumption for the natal kicks of neutron stars.  This increases the predictions for BHNS by a factor 3. The impact on BHBH systems is much smaller, This is because they mostly form from more massive cores which are already assumed to form with low kicks in our simulations.  As expected, the NSNS population is not affected.

%We note that the these two models best match the detection rates from LIGO (Broekgaarden et al.\ subm). Therefore we find the best matching models for the LIGO detection rates also produce the highest LISA detection rates.

\subsubsection{Stellar winds}
Mass loss in the form of stellar winds or eruptions are also a main incertainty. It affects by how much stars can grow in size and it affects the final core masses.

\citet{Broekgaarden+2021} provides us with model variations that consider variations in the mass loss by hot and luminous Wolf-Rayet stars.   We find that a reduction of the wind mass loss  has very little effect on our predictions.  It turns out that this is the consequence of several effects that cancel each other.

@@

In model \modWRHigh{} we instead \textit{increase} the efficiency of Wolf-Rayet winds by a factor of 5. We note that there is currently no observational support for such a strong increase in the stellar winds of Wolf-Rayet stars, but it is nevertheless still interesting to consider.  In this model the detection rate of BHBHs decreases by over a factor of 10 and BHNSs by over a factor of 2 whilst the NSNS rate is relatively unchanged. Increasing the efficiency of WR winds strongly decreases the final masses of DCOs. This means that many progenitors that would have formed LISA sources under our fiducial assumptions would not have enough mass to produce a DCO, or produce a NS instead of a BH (therefore this effect is stronger on BHBHs than NSNSs). Moreover, DCOs that are formed tend to be less massive and therefore less detectable.

>>>>

The efficiency of mass transfer, i.e.\ the fraction of mass lost by the donor through Roche lobe overflow that is accreted by the companion, is poorly constrained and is considered as one of the main uncertainties in binary evolution \citep[e.g.\ ][]{deMink+2007}. In our fiducial model A, we use a prescription in which the accretion rate onto stellar companions is regulated by their thermal timescale, i.e.\ the timescale on which a star can react to changes and restore thermal equilibrium \citep[see e.g.\ ][]{Schneider+2015}.

In models \modBetaLow{}-\modBetaHigh{}, we instead adopt a fixed value for the mass transfer efficiency, $\beta$, from $\beta=0.25$ up to 0.75, in cases of stable mass transfer onto a stellar companion. For accretion onto NS and BH we still assume that their accretion is limited to the Eddington rate.

@@

For the most massive progenitors, increasing $\beta$ leads to secondary stars that are so massive and luminous that they experience strong wind mass loss. This leads to further widening of the orbit and it limits the amount by which the secondaries expand. Both of these effects can prevent secondaries from filling their Roche lobe. This means that they cannot initiate the reverse interaction needed to shrink the binary system and eventually produce a detectable double compact object. We indeed see that increasing $\beta$ leads to a decrease of the expected number of detections for BHBHs and BHNSs, which originate from the most massive progenitors.

For lower mass progenitors, which primarily produce NSNS systems, we find the opposite: increasing $\beta$ leads to an \textit{increase} in the number of detectable NSNS systems. This is in part because the increased mass of the secondary stars allows stars that would have otherwise ended their life as a WD to instead become massive enough to form a NS. The same effect also allows some NS progenitors to become massive enough to BH progenitors, but this effect is less strong. Furthermore, the changes in secondary mass and orbital widening, affects the number of systems for which the reverse interaction successfully ejects the envelope and shrinks the orbit.

@@

The common-envelope phase constitutes the most challenging phase in the evolution of interacting binary systems \citep{Ivanova+2013}. The debate concerns the conditions required for the onset of a common envelope phase and, if a common envelope phase occurs, what the outcome is. Rapid population synthesis simulations such as ours approximate both questions in a crude way. We therefore consider several model variations.

\paragraph{The efficiency parameter $\alpha_{\rm CE}$} To estimate the outcome of a CE phase, we use a simple consideration of the binding energy and orbital energy  \citep{Webbink+1984}. Our fiducial model assumes a common-envelope efficiency parameter $\alpha_{\rm CE} = 1$ which can be interpreted as the case where all the energy liberated by shrinking the orbit is used in an optimal way to unbind the envelope \citep{Webbink+1984}. There have been many attempts to constrain this parameter using observations and more recently also using 3D simulations, but no consistent picture has emerged. Therefore we consider variations in this parameter.

@@

Increasing $\alpha_{\rm CE}$ makes common-envelope ejection more efficient or, in other words, less orbital shrinkage is needed to successfully eject the envelope. This has two consequences. (1) Firstly, a larger fraction of systems avoids merging during the CE phase. This increases the overall number of DCOs. (2) Secondly, the systems that survive the CE phase are wider, possibly too wide to become detectable as gravitational wave sources. So, while the first consequence favours the formation of DCOs, the second consequence disfavours the subset that are tight enough to be detected.

These two opposing effect result in a fine tuning situation. Only a very small subset of progenitor systems that have the right orbital parameters prior to the CE phase to successfully produce detectable systems. Changing  $\alpha_{\rm CE}$ moves and changes this window in the parameter space that successfully leads to the formation of detectable systems. How the number of the detectable systems changes depends on whether the relevant parameter space grows or shrinks and on how well the relevant part of the parameters space is populated. Fully unravelling these effects and how they interplay with the assumed star formation history is beyond our scope (and possibly not even of large relevance given the severe simplifications). We will limit the further discussions to simply stating the trends we observe.

@@

\paragraph{The ``optimistic'' CE treatment} We further explore a model variation introduced by \citet{Belczynski+2007} often referred to as the ``optimistic'' CE scenario. This variation (model \modOpt{}) relaxes our restriction that donor stars that are on the Hertzsprung cannot survive common-envelope events.

@@

\paragraph{Case BB mass transfer} For completeness and consistency with \citet{Broekgaarden+2021}, we have also included model variations that consider uncertainties in case BB mass transfer. This is a phase of mass transfer where the donor star has already lost its hydrogen envelope in a prior interaction, but fills its Roche lobe again as it expands during helium shell burning phase \citep[e.g.][]{Dewi+2003, Vigna-Gomez+2018}. This is of particular interest for the formation of NS stars, as their lower mass progenitors are swelling the most during this phase \citep[e.g.][and references therein]{Laplace+2020}. Nearly all NSNS systems from through case BB mass transfer

In model \modCaseBB{} we effectively exclude any systems that evolve through case BB mass transfer. As expected, this kills nearly all NSNS systems. We also find a heavy reduction of the BHNS systems by a factor 6. The BHBH systems are not affected, as expected, since case BB mass transfer does not play a role for high mass progenitors.

Finally, in model \modCaseBBOpt{}, we consider a situation where case BB mass transfer always leads to unstable mass transfer and initiates a CE phase in combination with the optimistic treatment for CE. This model leads to the highest predictions for the detections among all variations that we have considered.

@@

The formation of compact remnants and their associated natal kicks also constitute important uncertainties.

In model \modRapid{} we consider the so-called ``rapid'' remnant mass function by \citet{Fryer+2015} as an alternative to the ``delayed'' description used in our fiducial simulations. This affects the mass distribution (see Sec.~\ref{sec:lower_mass_gap}), but the effects on the number of detectable systems is modest for BHBH and BHNS, and negligible for NSNS.

The same is true for the impact of changing the assumed maximum neutron star mass, $m_{\rm NS, max}$ (\modNSLow{} and \modNSHigh{}). Lowering $m_{\rm NS, max}$ increases the number of detections involving BHs (since more stars form BHs instead of NSs) and vice versa, but has no significant effect on the number of NSNS detections since the vast majority are formed from low mass NSs.

We find that not implementing pair-instability supernovae (PISN) ot pulsational pair-instability supernovae (PPISN) in model \modNoPISN{} has no effect on the number of detections with LISA. This is because the average metallicity of the Milky Way is high enough than no progenitor retains enough mass to initiate a PISN or PPISN.

Decreasing the natal kicks for all core-collapse supernovae (models \modSigLow{}-\modSigLower{}) increases the detection rates for each DCO type since lower kicks result in fewer disrupted binaries and hence a more numerous detectable population. The BHNS and NSNS systems are strongly affected whilst the impact on BHBH systems is insignificant. The reason that BHBHs are relatively unaffected is that, in our models, the natal kicks for BHs are scaled down with the amount of mass that falls back. In the case of BHBHs, the black holes have very massive cores and thus low kicks.

In model \modNoBH{} we assume BHs form without any kick, while using our fiducial assumption for the natal kicks of neutron stars. This increases the predictions for BHNS by a factor 3 but the impact on BHBH systems is much smaller for the same reason as for models \modSigLow{}-\modSigLower{}. As expected, the NSNS population is not affected.

\subsubsection{Stellar winds}
Mass loss in the form of stellar winds or eruptions is also a main uncertainty. It affects by how much stars can grow in size and it affects the final core masses. \citet{Broekgaarden+2021} provides us with model variations that consider variations in the mass loss by hot and luminous Wolf-Rayet stars. We find that a reduction of the wind mass loss has very little effect on our predictions. This is the consequence of several effects that cancel each other.

@@

In model \modWRHigh{} we instead \textit{increase} the efficiency of Wolf-Rayet winds by a factor of 5. We note that though there is currently no observational support for such a strong increase in the stellar winds of Wolf-Rayet stars, it is nevertheless still interesting to consider. In this model the detection rate of BHBHs decreases by over a factor of 10 and BHNSs by over a factor of 2 whilst the NSNS rate is relatively unchanged. Increasing the efficiency of WR winds strongly decreases the final masses of DCOs. This means that many progenitors that would have formed LISA sources under our fiducial assumptions would not have enough mass to produce a DCO, or produce a NS instead of a BH (therefore this effect is stronger on BHBHs than NSNSs). Moreover, DCOs that are formed tend to be less massive and therefore less detectable.
